---
title: "Formatting entomological data with format_eht_data()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formatting entomological data with format_eht_data()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load-package}
library(AnophelesEntoTrials)
```

## Formatting of entomological data

Entomological hut trial (EHT) data often comes from various sources with inconsistent naming. Different teams may use:

- Different column names for the same variables (e.g., `UnfedAlive`, `UF.alive`, `u_alive`, `UA`)
- Different cases (e.g., `Treatment`, `treatment`, `TREATMENT`)
- Different treatment encodings (e.g., character labels vs. numeric codes)

The `format_eht_data()` function standartizes this data inconsistencies.

## What does format_eht_data() function do?

1. **Standardizes column names**: Maps common variations to standard names (UA, UD, FA, FD, treatment, etc.)
2. **Coerces data types**: Converts endpoints to numeric
3. **Encodes treatments**: Converts treatment labels to numeric encoding (0 = control, 1+ = treatments)
4. **Calculates variables**: Adds `fed`, `dead`, and `total` columns

## Example

```{r create-bad-data}
# A messy dataset with inconsistent naming conventions
messy_data <- data.frame(
  # Different naming for endpoints
  unfedalive = c(25, 30, 22, 28, 35, 15, 18, 12, 20, 14),
  UF.dead    = c(3, 2, 4, 3, 2, 8, 10, 12, 9, 11),
  fed_alive  = c(18, 15, 20, 17, 12, 10, 8, 6, 9, 7),
  F.dead     = c(2, 1, 3, 2, 1, 7, 9, 10, 8, 8),

  # Treatment as character with various labels
  nettype = c("Control", "Control", "Control", "Control", "Control",
              "IG2", "IG2", "IG2", "IG2", "IG2"),

  # Additional metadata
  hut_id = c(1, 2, 3, 4, 5, 1, 2, 3, 4, 5),
  night = c(1, 1, 1, 1, 1, 2, 2, 2, 2, 2)
)

print(messy_data)
str(messy_data)
```
Now let's format this data:

```{r format-data}
# Apply format_eht_data()
clean_data <- format_eht_data(messy_data, control_value = "Control")

print(clean_data)
str(clean_data)
```

## Handling Different Naming Conventions

The `standardize_column_names()` function (called internally) recognizes many common variations:

| Standard Name | Recognized Variations |
|--------------|----------------------|
| UA | `unfedalive`, `UF.alive`, `u_alive`, `unfed_alive`, `gaufalive24` |
| UD | `unfeddead`, `UF.dead`, `u_dead`, `unfed_dead`, `gaufdead24` |
| FA | `fedalive`, `F.alive`, `f_alive`, `fed_alive`, `gafedalive24` |
| FD | `feddead`, `F.dead`, `f_dead`, `fed_dead`, `gafeddead24` |
| treatment | `treat`, `nettype`, `treatment_type` |
| vector_control_product | `insecticide_name`, `insecticide`, `net_type`, `lntype` |
| wash_status | `washed`, `condition`, `wash` |

## Specifying the control group

By default, `format_eht_data()` attempts to auto-detect the control group by looking for:
1. A treatment value of `0`
2. A treatment label containing "control" (case-insensitive)

If neither is found, it uses the first unique value and issues a warning. You can explicitly specify the control:

```{r specify-control}
# Data with non-obvious control labeling
data_odd_labels <- data.frame(
  UA = c(30, 28, 15, 12),
  UD = c(2, 3, 10, 11),
  FA = c(15, 17, 8, 6),
  FD = c(1, 2, 9, 10),
  treatment = c("Untreated_Net", "Untreated_Net", "PBO_Net", "PBO_Net")
)

# Explicitly specify control
formatted <- format_eht_data(data_odd_labels, control_value = "Untreated_Net")
print(formatted[, c("treatment", "UA", "UD", "FA", "FD")])
```

## Skipping Derived Variables

If you don't need the derived variables (`fed`, `dead`, `total`), you can skip their calculation:

```{r skip-derived}
minimal_format <- format_eht_data(messy_data,
                                   calculate_derived = FALSE,
                                   control_value = "Control")
names(minimal_format)
```

## Integration with the analysis pipeline

```{r pipeline-example, eval = FALSE}
# 1. Import raw data
raw_data <- import_ento_data("my_trial_data.csv")

# 2. Format the data
formatted_data <- format_eht_data(raw_data, control_value = "Control")

# 3. Fit the model
fit <- fit_ento_model(
  data = formatted_data,
  model_type = "basic",
  iter = 6000,
  warmup = 3000,
  chains = 3
)
```