
---
title: "Workflow: Using Custom Stan Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow: Using Custom Stan Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```
This vignette demonstrates how to use custom Stan models with the `AnophelesEntoTrials` package. This is useful when you need:

- Different model structure (e.g., additional covariates)
- Different priors
- Custom parameters
- Novel outcome distributions

## Overview

The package supports custom models through three mechanisms:

1. **Custom Stan file**: Provide path to your own `.stan` file
2. **Custom Stan data**: Provide your own data list matching your model's requirements
3. **Custom parameters**: Specify which parameters to extract
4. **Custom initialization**: Provide your own initialization function

## Example 1: Simple Custom Model with Additional Parameter

Let's create a custom model that includes an additional parameter for net age effect.

### Step 1: Write Custom Stan Model

Create a file `custom_model_net_age.stan`:

```{stan, output.var = "custom_model", eval = FALSE}
// Custom EHT model with net age covariate
functions {
  real P_UA(real alpha, real mu) {
    return exp(-(alpha + mu));
  }
  real P_F(real alpha, real mu) {
    return (1 - P_UA(alpha, mu)) * alpha / (alpha + mu);
  }
  real P_UD(real alpha, real mu) {
    return (1 - P_UA(alpha, mu)) * mu / (alpha + mu);
  }
  real P_FA(real alpha, real mu, real PB) {
    return P_F(alpha, mu) * PB;
  }
  real P_FD(real alpha, real mu, real PB) {
    return P_F(alpha, mu) * (1 - PB);
  }
}

data {
  int<lower=0> N;                // number of observations
  int<lower=0> tr;               // number of treatments
  int treat[N];                  // treatment index
  int<lower=0> y[N, 4];          // outcomes (UA, UD, FA, FD)
  vector[N] net_age;             // net age in months (NEW!)
  int<lower=0, upper=1> use_likelihood;
}

parameters {
  vector<lower=0, upper=1>[tr+1] InitialPostprandialkillingEfficacy;
  vector<lower=0>[tr+1] KillingDuringHostSeeking;
  vector<lower=0, upper=1>[tr+1] InitialRepellencyRate;

  // NEW: Net age decay parameter
  real<lower=0> net_age_decay;

  real a;
  real m;
  real b;
}

transformed parameters {
  real<lower=0> alpha_0 = exp(a);
  real<lower=0> mu_0 = exp(m);
  real<lower=0, upper=1> pB_0 = inv_logit(b);
}

model {
  matrix[4, N] theta;

  // Priors
  target += uniform_lpdf(InitialPostprandialkillingEfficacy | 0, 1);
  target += lognormal_lpdf(1 - InitialRepellencyRate | 0, 5);
  target += lognormal_lpdf(KillingDuringHostSeeking | 0, 5);

  // Prior for net age decay (half-normal)
  target += normal_lpdf(net_age_decay | 0, 0.1) - normal_lccdf(0 | 0, 0.1);

  target += normal_lpdf(a | 0, 6);
  target += normal_lpdf(m | 0, 6);
  target += logistic_lpdf(b | 0, 1);

  // Likelihood
  for (i in 1:N) {
    // Calculate decay factor based on net age
    real decay_factor = exp(-net_age_decay * net_age[i]);

    if (treat[i] == 0) {
      theta[1, i] = P_UA(alpha_0, mu_0);
      theta[2, i] = P_UD(alpha_0, mu_0);
      theta[3, i] = P_FA(alpha_0, mu_0, pB_0);
      theta[4, i] = P_FD(alpha_0, mu_0, pB_0);
    } else {
      // Apply efficacy with decay
      real eff_repellency = InitialRepellencyRate[treat[i]+1] * decay_factor;
      real eff_killing = KillingDuringHostSeeking[treat[i]+1] * decay_factor;
      real eff_postprandial = InitialPostprandialkillingEfficacy[treat[i]+1] * decay_factor;

      real alpha_i = (1 - eff_repellency) * alpha_0;
      real mu_i = mu_0 + eff_killing * alpha_0;
      real pBi = pB_0 * (1 - eff_postprandial);

      theta[1, i] = P_UA(alpha_i, mu_i);
      theta[2, i] = P_UD(alpha_i, mu_i);
      theta[3, i] = P_FA(alpha_i, mu_i, pBi);
      theta[4, i] = P_FD(alpha_i, mu_i, pBi);
    }

    if (use_likelihood == 1)
      target += multinomial_lpmf(y[i, ] | theta[, i]);
  }
}

generated quantities {
  // Compute half-life of net effectiveness
  real net_halflife = log(2) / net_age_decay;

  // Other derived quantities as needed
}
```

### Step 2: Prepare Custom Data

```{r prepare-custom-data}
library(AnophelesEntoTrials)

# Create data with net age
custom_data <- data.frame(
  UA = sample(10:50, 100, replace = TRUE),
  UD = sample(0:10, 100, replace = TRUE),
  FA = sample(5:30, 100, replace = TRUE),
  FD = sample(0:5, 100, replace = TRUE),
  treatment = rep(c(0, 1), each = 50),
  net_age = c(rep(0, 50), runif(50, 0, 36))  # Net age in months
)

# Prepare Stan data manually
custom_stan_data <- list(
  N = nrow(custom_data),
  tr = length(unique(custom_data$treatment)) - 1,
  treat = as.integer(custom_data$treatment),
  y = as.matrix(custom_data[, c("UA", "UD", "FA", "FD")]),
  net_age = custom_data$net_age,  # Custom variable!
  use_likelihood = 1
)
```

### Step 3: Define Custom Initialization

```{r custom-init}
# Custom initialization function
custom_init <- function() {
  list(
    InitialPostprandialkillingEfficacy = runif(2, 0.2, 0.8),
    KillingDuringHostSeeking = runif(2, 0.1, 0.5),
    InitialRepellencyRate = runif(2, 0.2, 0.8),
    net_age_decay = runif(1, 0.01, 0.05),  # Custom parameter init
    a = rnorm(1, -2, 0.5),
    m = rnorm(1, -2, 0.5),
    b = rnorm(1, 0, 1)
  )
}
```

### Step 4: Fit Custom Model

```{r fit-custom-model}
# Fit with custom model
fit_custom <- fit_ento_model(
  data = custom_data,
  custom_model = "path/to/custom_model_net_age.stan",
  stan_data = custom_stan_data,
  pars = c(
    "InitialPostprandialkillingEfficacy",
    "KillingDuringHostSeeking",
    "InitialRepellencyRate",
    "net_age_decay",      # Custom parameter!
    "net_halflife",       # Derived quantity
    "alpha_0",
    "mu_0"
  ),
  init = custom_init,
  iter = 6000,
  warmup = 3000,
  chains = 3,
  seed = 123
)
```

### Step 5: Extract Custom Results

```{r extract-custom-results}
# Extract summary for all parameters including custom ones
summary_custom <- extract_stan_summary(
  fit_custom,
  pars = c("net_age_decay", "net_halflife",
           "InitialPostprandialkillingEfficacy",
           "InitialRepellencyRate",
           "KillingDuringHostSeeking")
)

print(summary_custom)

# Or extract all parameters directly from rstan
library(rstan)
print(fit_custom, pars = c("net_age_decay", "net_halflife"))
```

## Example 2: Model with Different Outcome Structure

Here's how to create a model with a different outcome structure (e.g., binary outcomes).

### Custom Binary Outcome Model

```{stan, output.var = "binary_model", eval = FALSE}
// Binary outcome model: alive vs dead
data {
  int<lower=0> N;
  int<lower=0> n_alive[N];    // Number alive
  int<lower=0> n_total[N];    // Total mosquitoes
  int treat[N];
  int<lower=0> tr;
}

parameters {
  real<lower=0, upper=1> baseline_survival;
  vector<lower=0, upper=1>[tr] mortality_effect;
}

model {
  // Priors
  baseline_survival ~ beta(5, 2);
  mortality_effect ~ beta(2, 2);

  // Likelihood
  for (i in 1:N) {
    real p_alive;
    if (treat[i] == 0) {
      p_alive = baseline_survival;
    } else {
      p_alive = baseline_survival * (1 - mortality_effect[treat[i]]);
    }
    n_alive[i] ~ binomial(n_total[i], p_alive);
  }
}
```

### Fitting the Binary Model

```{r fit-binary-model}
# Prepare data for binary model
binary_data <- data.frame(
  n_alive = c(rep(45, 50), rep(30, 50)),
  n_total = rep(50, 100),
  treatment = rep(c(0, 1), each = 50)
)

binary_stan_data <- list(
  N = nrow(binary_data),
  n_alive = binary_data$n_alive,
  n_total = binary_data$n_total,
  treat = as.integer(binary_data$treatment),
  tr = 1
)

# Fit
fit_binary <- fit_ento_model(
  data = binary_data,
  custom_model = "path/to/binary_model.stan",
  stan_data = binary_stan_data,
  pars = c("baseline_survival", "mortality_effect"),
  init = function() {
    list(
      baseline_survival = runif(1, 0.7, 0.9),
      mortality_effect = runif(1, 0.1, 0.5)
    )
  }
)
```

## Example 3: Extending Predefined Models

You can also extend the predefined models by copying and modifying them.

### Get Path to Predefined Model

```{r get-model-path}
# Find the package's Stan model file
basic_model_path <- system.file("stan",
                                "mosq_foraging_mcmc_full_fit_priorpred.stan",
                                package = "AnophelesEntoTrials")

# Read and modify
model_code <- readLines(basic_model_path)
head(model_code, 30)

# Copy to your working directory for modification
file.copy(basic_model_path, "my_modified_model.stan")
```

## Best Practices for Custom Models

### 1. Data Validation

Always validate your custom data matches model requirements:

```{r validate-data}
validate_custom_data <- function(stan_data, required_vars) {
  missing <- setdiff(required_vars, names(stan_data))
  if (length(missing) > 0) {
    stop("Missing required variables: ", paste(missing, collapse = ", "))
  }

  # Check dimensions
  if ("y" %in% names(stan_data)) {
    if (nrow(stan_data$y) != stan_data$N) {
      stop("y matrix rows must equal N")
    }
  }

  message("Data validation passed!")
  invisible(TRUE)
}

# Use before fitting
validate_custom_data(
  custom_stan_data,
  required_vars = c("N", "tr", "treat", "y", "net_age", "use_likelihood")
)
```

### 2. Prior Predictive Checks

Test your model with prior predictive checks before fitting real data:

```{r prior-predictive}
# Set use_likelihood = 0 for prior predictive
prior_pred_data <- custom_stan_data
prior_pred_data$use_likelihood <- 0

fit_prior <- fit_ento_model(
  data = custom_data,
  custom_model = "path/to/custom_model.stan",
  stan_data = prior_pred_data,
  iter = 2000,
  chains = 2
)
```

### 3. Model Comparison

Compare custom models with predefined ones:

```{r model-comparison}
# Fit basic model
fit_basic <- fit_ento_model(
  data = custom_data[, c("UA", "UD", "FA", "FD", "treatment")],
  model_type = "basic"
)

# Fit custom model
fit_custom <- fit_ento_model(
  data = custom_data,
  custom_model = "path/to/custom_model.stan",
  stan_data = custom_stan_data,
  pars = c("InitialPostprandialkillingEfficacy",
           "KillingDuringHostSeeking",
           "InitialRepellencyRate")
)

# Compare using LOO-CV (requires loo package)
library(loo)
loo_basic <- loo(fit_basic)
loo_custom <- loo(fit_custom)
loo_compare(loo_basic, loo_custom)
```

## Troubleshooting Custom Models

### Common Issues

1. **Parameter dimension mismatch**: Ensure initialization returns vectors of correct length

```{r fix-dimension}
# Check number of treatments
n_treat <- length(unique(custom_data$treatment))

# Initialization must match
init_fn <- function() {
  list(
    # Vector parameters need length = n_treat
    InitialPostprandialkillingEfficacy = runif(n_treat, 0.1, 0.9)
  )
}
```

2. **Data type errors**: Stan is strict about integer vs real

```{r fix-types}
# Ensure integers where needed
stan_data$N <- as.integer(stan_data$N)
stan_data$y <- matrix(as.integer(stan_data$y), ncol = 4)
stan_data$treat <- as.integer(stan_data$treat)
```

3. **Missing generated quantities**: Add them to pars

```{r include-generated}
fit <- fit_ento_model(
  # ... other args
  pars = c("alpha_0", "mu_0", "net_halflife")  # Include derived quantities
)
```

## Summary

Key steps for custom models:

1. Write your Stan model file
2. Prepare matching `stan_data` list
3. Define custom `init` function
4. Specify `pars` to extract
5. Validate and test before full fitting

See `?fit_ento_model` for all available options.
