---
title: "Workflow: Using Predefined Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow: Using Predefined Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette demonstrates the complete workflow for analyzing entomological trials data using Bayesian models in the `AnophelesEntoTrials` package.

## Overview

(for more details about parameters and model details, see below)
The package provides three predefined Stan models for different experimental designs:

1. **Basic model** (`"basic"`): Standard multinomial model for typical EHT data
2. **Hierarchical decay model** (`"hierarchical_decay"`): For longitudinal data with temporal decay
3. **Semifield model** (`"semifield"`): For semifield experiments with time-to-event outcomes

## Step 1: Load the Package

```{r load-package}
library(AnophelesEntoTrials)

# See available models
list_available_models()
```

## Step 2: Prepare Your Data

### Data Requirements

Each model has specific data requirements. Check what's needed:
```{r check-requirements}
# View model details including required columns
models_info <- list_available_models(details = TRUE)
print(models_info)
```

### Basic Model Data Format

The basic model requires:
- `UA`: Unfed Alive count
- `UD`: Unfed Dead count
- `FA`: Fed Alive count
- `FD`: Fed Dead count
- `treatment`: Treatment indicator (0 = control, 1+ = treatments)

```{r prepare-basic-data}
# Example: Import data from CSV
eht_data <- import_ento_data("path/to/your/data.csv")
# Or define your required columns
eht_data_treat <- import_ento_data("path/to/your/data.csv",
                                   required_vars = c("deadfed", "alivefed", "alive"),
                                   treatment_var = "Intervention")

# ------------- Simulated data example ----------
# ------------- Use it for testing purposes -----
set.seed(42)

# True parameters for simulation
mu_0 <- 0.14      # baseline mortality rate
alpha_0 <- 0.15   # baseline attack rate
pB_0 <- 0.92      # baseline survival probability post-feeding (blood-meal survival)

# Treatment effects
pi <- 0.69        # repellency (69% deterrent effect)
kappa <- 3.8      # pre-prandial killing multiplier
xi <- 0.52        # post-prandial killing (52% killed after feeding)

# Probability function for mosquito outcomes
return_proba <- function(alpha, mu, pc) {
  c(
    exp(-alpha - mu),                                    # UA: unfed alive
    (1 - exp(-alpha - mu)) * mu / (alpha + mu),          # UD: unfed dead
    (1 - exp(-alpha - mu)) * alpha / (alpha + mu) * pc,  # FA: fed alive
    (1 - exp(-alpha - mu)) * alpha / (alpha + mu) * (1 - pc)  # FD: fed dead
  )
}

# Calculate probabilities for control and treatment
proba_control <- return_proba(alpha_0, mu_0, pB_0)
proba_treatment <- return_proba(
  alpha_0 * (1 - pi),           # reduced attack rate due to repellency
  mu_0 + alpha_0 * kappa,       # increased mortality
  pB_0 * (1 - xi)               # reduced survival post-feeding
)

# Generate simulated data
n_samples <- 100

sample_control <- data.frame(t(rmultinom(n_samples, 10, proba_control)))
sample_treatment <- data.frame(t(rmultinom(n_samples, 10, proba_treatment)))

# Combine into dataset
simulated_data <- rbind(
  data.frame(sample_control, treatment = 0, insecticide_name = "Control"),
  data.frame(sample_treatment, treatment = 1, insecticide_name = "NewITN")
)
names(simulated_data)[1:4] <- c("UA", "UD", "FA", "FD")

# Format the data (adds derived columns)
eht_data <- format_eht_data(simulated_data)
head(eht_data)
```

## Step 3: Fit the Model

### Basic Model

```{r fit-basic-model}
# Fit with default settings
fit_basic <- fit_ento_model(
  data = eht_data,
  model_type = "basic",
  iter = 6000,
  warmup = 3000,
  chains = 3,
  seed = 123
)

# Or with adjusted MCMC settings for better convergence
fit_basic <- fit_ento_model(
  data = eht_data,
  model_type = "basic",
  iter = 10000,
  warmup = 5000,
  chains = 4,
  control = list(adapt_delta = 0.95, max_treedepth = 12),
  seed = 123
)
```

### Hierarchical Decay Model

For longitudinal data with temporal decay effects:

```{r fit-decay-model}
# Prepare longitudinal data (requires additional columns)
longitudinal_data <- data.frame(
  UA = sample(10:50, 200, replace = TRUE),
  UD = sample(0:10, 200, replace = TRUE),
  FA = sample(5:30, 200, replace = TRUE),
  FD = sample(0:5, 200, replace = TRUE),
  treatment = rep(c(0, 1), each = 100),
  day = rep(1:10, 20),        # Day index (1 to D)
  datej = rep(1:100, 2),      # Days since treatment start
  fed = NA,                    # Will be calculated
  total = NA                   # Will be calculated
)
longitudinal_data$fed <- longitudinal_data$FA + longitudinal_data$FD
longitudinal_data$total <- rowSums(longitudinal_data[, c("UA", "UD", "FA", "FD")])

# Fit hierarchical decay model
fit_decay <- fit_ento_model(
  data = longitudinal_data,
  model_type = "hierarchical_decay",
  iter = 8000,
  warmup = 4000,
  chains = 3,
  prior_settings = list(
    priorsigma_mean_logrates = 6,
    hierarchy = 1
  ),
  seed = 123
)
```

## Step 4: Extract and Interpret Results

### Summary Statistics

```{r extract-summary}
# Extract posterior summaries with 95% credible intervals
summary_results <- extract_stan_summary(
  fit_basic,
  data = eht_data,
  probs = c(0.025, 0.975)
)

print(summary_results)
```

### posteriorMax Estimates

```{r extract-map}
# Get posteriorMax estimates (single best estimate)
posteriorMax <- extract_posteriorMax(fit_basic, data = eht_data)
print(posteriorMax)
```

### Full Posterior Samples

```{r extract-posteriors}
# Extract subset of posterior samples for further analysis
posterior_samples <- extract_stan_posteriors(
  fit_basic,
  data = eht_data,
  n_samples = 500  # Random subset of 500 samples
)

# Visualize posterior distribution
library(ggplot2)

ggplot(posterior_samples, aes(x = value, fill = param)) +
geom_density(alpha = 0.5) +
  facet_wrap(~ param, scales = "free") +
  theme_minimal() +
  labs(title = "Posterior Distributions of Intervention Effects")
```

## Step 5: Save and Load Results

```{r save-results}
# Save during fitting
fit <- fit_ento_model(
  data = eht_data,
  model_type = "basic",
  output_path = "results/my_fit.rds"
)

# Or save manually
saveRDS(fit_basic, "results/fit_basic.rds")

# Load later
fit_loaded <- readRDS("results/fit_basic.rds")
```

## Interpreting Parameters

The basic model estimates the following intervention parameters:

| Parameter | Symbol | Interpretation | Range |
|-----------|--------|----------------|-------|
| InitialRepellencyRate | π | Proportion of mosquitoes deterred from entering | 0-1 |
| KillingDuringHostSeeking | κ | Additional mortality during approach (multiplier) | ≥0 |
| InitialPostprandialkillingEfficacy | ξ | Proportion killed after blood-feeding | 0-1 |
| InitialPreprandialkillingEfficacy | - | Derived: overall pre-feeding killing | 0-1 |

Baseline parameters:
- `alpha_0`: Baseline attack rate (exp(a) in Stan)
- `mu_0`: Baseline mortality rate (exp(m) in Stan)
- `pB_0`: Baseline probability of surviving blood-feeding (inv_logit(b) in Stan)

## Understanding Parameter Initialization

MCMC sampling requires starting values (initial values) for all parameters. The package provides automatic initialization, but understanding how it works helps with troubleshooting and customization.

### How Initialization Works

When you call `fit_ento_model()` with `init = "auto"` (the default), the package:

1. **Creates an init function** using `generate_inits_basic()`, `generate_inits_decay()`, or `generate_inits_semifield()` depending on your model type
2. **Stan calls this function once per chain** to get different starting points
3. **Each chain starts from a different random location** in parameter space

```{r init-example}
# See what the init function produces
library(AnophelesEntoTrials)

# Create init function for 2 treatment groups (control + 1 intervention)
init_fn <- generate_inits_basic(nb_treat = 2)

# Call it to see one set of initial values
init_fn()
```

### Why Multiple Chains with Different Inits?

Different chains starting from different points should converge to the same posterior distribution if:
- The model is well-specified
- There's enough data
- The sampler has run long enough

If chains don't converge (Rhat > 1.1), it may indicate:
- Insufficient iterations
- Model identifiability issues
- Need for better initialization

### Customizing Initialization

You can provide your own init function for more control:

```{r custom-init}
# Custom init function with specific starting values
my_inits <- function() {
  list(
    InitialPostprandialkillingEfficacy = c(0.1, 0.5),  # control, treatment
    KillingDuringHostSeeking = c(0.1, 0.5),
    InitialRepellencyRate = c(0.1, 0.5),
    a = -2,  # log(alpha_0) ≈ log(0.14) = -2
    m = -2,  # log(mu_0) ≈ log(0.14) = -2
    b = 2    # logit(pB_0) ≈ logit(0.88) = 2
  )
}

# Use custom inits
fit <- fit_ento_model(
  data = eht_data,
  model_type = "basic",
  init = my_inits
)
```

### Parameter Scales in Stan

The Stan models use transformed scales for baseline parameters:

| R Parameter | Stan Parameter | Transformation | Typical Range |
|-------------|----------------|----------------|---------------|
| `alpha_0` | `a` | `alpha_0 = exp(a)` | a ∈ (-4, 0) |
| `mu_0` | `m` | `mu_0 = exp(m)` | m ∈ (-4, 0) |
| `pB_0` | `b` | `pB_0 = inv_logit(b)` | b ∈ (-2, 4) |

When providing custom inits, specify values on the **transformed scale** (a, m, b) not the natural scale.

## Comparing Multiple Treatments

When you have multiple treatments in your data:

```{r multiple-treatments}
# Data with multiple treatments
multi_treatment_data <- data.frame(
  UA = c(rep(30, 50), rep(25, 50), rep(20, 50)),
  UD = c(rep(5, 50), rep(8, 50), rep(12, 50)),
  FA = c(rep(15, 50), rep(12, 50), rep(8, 50)),
  FD = c(rep(2, 50), rep(5, 50), rep(8, 50)),
  treatment = rep(c(0, 1, 2), each = 50),
  insecticide_name = rep(c("Control", "Standard_ITN", "New_ITN"), each = 50)
)

multi_treatment_data <- format_eht_data(multi_treatment_data)

# Fit model
fit_multi <- fit_ento_model(
  data = multi_treatment_data,
  model_type = "basic"
)

# Compare treatments
summary_multi <- extract_stan_summary(fit_multi, data = multi_treatment_data)
print(summary_multi)
```

## Troubleshooting

### Divergent Transitions

If you see warnings about divergent transitions:

```{r troubleshoot-divergent}
# Increase adapt_delta
fit <- fit_ento_model(
  data = eht_data,
  model_type = "basic",
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```

### Slow Convergence

If Rhat values are > 1.1:

```{r troubleshoot-convergence}
# Increase iterations and chains
fit <- fit_ento_model(
  data = eht_data,
  model_type = "basic",
  iter = 20000,
  warmup = 10000,
  chains = 4
)
```

## Next Steps

- See `vignette("workflow-custom-models")` for using custom Stan models
- See `?extract_stan_summary` for more extraction options
- See `?list_available_models` for model details
